Analysing the effects of COVID-19 on Crime in the United States

### 2022-05-01 outputs/paper/paper.Rmd
Added complete results section with analysis of Granger tests

\newpage
# Methods

@@

\newpage
# Results 
Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.

@@

Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.

@@

combined_diff2 <- merge(vehicle_diff[paste(common_time[1],
                                            common_time[length(common_time)],
                                            sep = "/")],
                       CH_res_ts[paste(common_time[1],
                                            common_time[length(common_time)],
                                            sep = "/")])
optimal_vehicle <- VARselect(na.omit(combined_diff2)[,c(1,2)], type = 'none', lag.max = 10)
VAR_vehicle <- VAR(y=as.ts(na.omit(combined_diff2)[,c(1,2)]), p=optimal_vehicle$selection[1])
# 
# causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
# causality(VAR_vehicle, cause = "residuals")
```
```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
par(mfrow = c(1,2))
lags = c(1:16)

# criminal damange
# only significant one is from covid to crime
irf_criminal_2 <- irf(VAR_criminal, 
                         impulse = "residuals", 
                         response = "criminal", 
                         n.ahead = 15,
                         ortho = F)

i1 <- ggplot(irf_criminal_2_gg, aes(x = lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new COVID-19 case")+
  ylab("Criminal Damage")+
  theme(text = element_text(size=10), 
        plot.title = element_text(hjust = 0.5))
i1

autoplot(f_criminal$forecast$criminal) +
  ggtitle("Prediction on how many more criminal damage cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
par(mfrow = c(1,2))
lags = c(1:16)

@@

## Los Angeles
Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
lags = c(1:16)

@@

# battery
# criem significant to covid
irf_battery_1 <- irf(VAR_battery,
                     impulse = "battery",
                     response = "residuals",
                     n.ahead = 15)

# html plot for battery
irf_battery_1_gg <- data.frame(
  irf_battery_1$irf$battery[,1],
  irf_battery_1$Lower$battery[,1],
  irf_battery_1$Upper$battery[,1]
)

colnames(irf_battery_1_gg) <- c("mean", "lower", "upper")

irf_battery_1_plot <- ggplot(irf_battery_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new Battery case")+
  ylab("COVID-19")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

irf_battery_1_plot

f_battery <- forecast::forecast(VAR_battery)

autoplot(f_battery$forecast$battery) +
  ggtitle("Prediction on how many more Battery cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
lags = c(1:16)

# battery
# criem significant to covid
irf_burglary_1 <- irf(VAR_burglary,
                     impulse = "burglary",
                     response = "residuals",
                     n.ahead = 15)

# html plot for burglary
irf_burglary_1_gg <- data.frame(
  irf_burglary_1$irf$burglary[,1],
  irf_burglary_1$Lower$burglary[,1],
  irf_burglary_1$Upper$burglary[,1]
)

irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new burglary case")+
  ylab("COVID-19")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

autoplot(f_burglary$forecast$burglary) +
  ggtitle("Prediction on how many more burglary cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}

@@

# general offense
# irf
irf_offense_1 <- irf(VAR_offense,
                     impulse = "offense",
                     response = "residuals",
                     n.ahead = 15)
# ggplot
irf_offense_1_gg <- data.frame(
  irf_offense_1$irf$offense[,1],
  irf_offense_1$Lower$offense[,1],
  irf_offense_1$Upper$offense[,1]
  )

colnames(irf_offense_1_gg) <- c("mean", "lower", "upper")

irf_offense_1_plot <- ggplot(irf_offense_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a New Other Offense Case")+
  ylab("COVID-19")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

irf_offense_1_plot

autoplot(f_offense$forecast$offense) +
  ggtitle("Prediction on how many more other offense cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}

# theft
# irf
irf_theft_1 <- irf(VAR_thefts,
                   impulse = "thefts",
                   response = "residuals",
                   n.ahead = 15)

# ggplot
irf_theft_1_gg <- data.frame(
  irf_theft_1$irf$thefts[,1],
  irf_theft_1$Lower$thefts[,1],
  irf_theft_1$Upper$thefts[,1]
)

@@

irf_theft_1_plot <- ggplot(irf_theft_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new Theft case")+
  ylab("COVID-19")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

@@

## Seattle 
Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.

@@

```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
causality(VAR_larceny, cause = colnames(combined_diff)[1])
causality(VAR_larceny, cause = "residuals")
causality(VAR_burglary, cause = colnames(combined_diff)[2])
causality(VAR_burglary, cause = "residuals")
causality(VAR_vandalism, cause = colnames(combined_diff)[3])
causality(VAR_vandalism, cause = "residuals")
causality(VAR_vehicle_theft, cause = colnames(combined_diff)[4])
causality(VAR_vehicle_theft, cause = "residuals")
causality(VAR_theft_fromvehicle, cause = colnames(combined_diff)[5])
causality(VAR_theft_fromvehicle, cause = "residuals")

@@

```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic. Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
lags = c(1:16)

@@

irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("How many more burglary cases per day there will be 
          after 1 confirmed covid19 case") +
  xlab("Number of days after a confimed COVID-19 case")+
  ylab("Number of bulglary cases")

@@

>>>>

Next, we take a look at the modelling approaches employed to further verify correlations between COVID-19 and crime data.

# Methods

@@

# Results 
We now employ the methods outlined in the previous section to narrow down on the crimes that are the most significantly correlated with COVID-19 activity in the respective cities. For each city, the impulse response alongside the model predictions are visualized for the crimes that are most significantly associated with COVID-19.

@@

Based on the fitted VAR models for top crimes in Chicago, we find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as criminal damage ($p=0.8271$), and battery ($p=0.63$). Whereas, a statistically significant instantaneous correlation was found between COVID-19 cases impacting assault ($p=0.036$) and a significant Granger-causality with the cases impacting vehicle theft ($p = 0.00022$). The impulse response of a new case to these crimes and the model predictions are outlined in Figures \@ref(fig:r11) & \@ref(fig:r12).

@@

combined_diff2 <- merge(vehicle_diff[paste(common_time[1],
                                            common_time[length(common_time)],
                                            sep = "/")],
                       CH_res_ts[paste(common_time[1],
                                            common_time[length(common_time)],
                                            sep = "/")])
optimal_vehicle <- VARselect(na.omit(combined_diff2)[,c(1,2)], type = 'none', lag.max = 10)
VAR_vehicle <- VAR(y=as.ts(na.omit(combined_diff2)[,c(1,2)]), p=optimal_vehicle$selection[1])
# 
causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
causality(VAR_vehicle, cause = "residuals")
```

```{r r11, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on assault (left), and VAR model prediction for assault (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
par(mfrow = c(1,2))
lags = c(1:16)

# criminal damange
# only significant one is from covid to crime
irf_assault_2 <- irf(VAR_assault, 
                         impulse = "residuals", 
                         response = "assault", 
                         n.ahead = 15,
                         ortho = F)

# ggplot version of it.
irf_assault_2_gg <- data.frame(
  irf_assault_2$irf$residuals[,1],
  irf_assault_2$Lower$residuals[,1],
  irf_assault_2$Upper$residuals[,1]
)

colnames(irf_assault_2_gg) <- c("mean", "lower", "upper")

i1 <- ggplot(irf_assault_2_gg, aes(x = lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new COVID-19 case")+
  ylab("Assualt")+
  theme(text = element_text(size=10), 
        plot.title = element_text(hjust = 0.5))
i1

f_assault <- forecast::forecast(VAR_assault)

autoplot(f_assault$forecast$assault) +
  ggtitle("Prediction on how many more assault cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

```{r r12, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vehicle thefts (left), and VAR model prediction for vehicle thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
par(mfrow = c(1,2))
lags = c(1:16)

@@

When it comes to Los Angeles, only vandalism stood out based on the fitted VAR models. We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as battery ($p=0.67$), theft from vehicle ($p=0.9$), vehicle theft ($p=0.81$),  and burglary ($p=0.26$). Whereas, a statistically significant Granger causality was found between COVID-19 cases impacting vandalism ($p=0.003$). The impulse response of a new case to vandalism and the model predictions are outlined in Figure \@ref(fig:r21).

@@

```{r r21, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vandalism (left), and VAR model prediction for vandalism (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
# use car theft and criminal damage
lags = c(1:16)

@@

# battery
# criem significant to covid
irf_vandalism_1 <- irf(VAR_vandalism,
                     impulse = "residuals",
                     response = "vandalism",
                     n.ahead = 15)

# html plot for vandalism
irf_vandalism_1_gg <- data.frame(
  irf_vandalism_1$irf$residuals[,1],
  irf_vandalism_1$Lower$residuals[,1],
  irf_vandalism_1$Upper$residuals[,1]
)

colnames(irf_vandalism_1_gg) <- c("mean", "lower", "upper")

irf_vandalism_1_plot <- ggplot(irf_vandalism_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new COVID-19 case")+
  ylab("Vandalism")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

irf_vandalism_1_plot

f_vandalism <- forecast::forecast(VAR_vandalism)

autoplot(f_vandalism$forecast$vandalism) +
  ggtitle("Prediction on how many more vandalism cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

@@

Coming to Philadelphia, only vehicle and general thefts stood out based on the fitted VAR models. We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as offense ($p=0.42$) and assault ($p=0.97$). On the contrary, a close statistically significant Granger causality was found between COVID-19 cases impacting vehicle theft ($p=0.0009$) and an instantaneous causality was found with thefts ($0.005$). The impulse response of a new case on these crimes and the model predictions are outlined in Figures \@ref(fig:r31) and \@ref(fig:r32).

@@

```{r r31, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vehicle thefts (left), and VAR model prediction for vehicle thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}

@@

# general vehicle theft
# irf
irf_vehicle_1 <- irf(VAR_vehicle,
                     impulse = "residuals",
                     response = "vehicle",
                     n.ahead = 15)
# ggplot
irf_vehicle_1_gg <- data.frame(
  irf_vehicle_1$irf$residuals[,1],
  irf_vehicle_1$Lower$residuals[,1],
  irf_vehicle_1$Upper$residuals[,1]
  )

colnames(irf_vehicle_1_gg) <- c("mean", "lower", "upper")

irf_vehicle_1_plot <- ggplot(irf_vehicle_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a New COVID-19 Case")+
  ylab("Vehicle Thefts")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

irf_vehicle_1_plot

f_vehicle <- forecast::forecast(VAR_vehicle)

autoplot(f_vehicle$forecast$vehicle) +
  ggtitle("Prediction on how many more vehicle theft cases compared to yesterday") +
  theme_minimal() +
  ylab("Day-to-day change") +
  xlab(paste("Number of days since", common_time[1]))

@@

```{r r32, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on thefts (left), and VAR model prediction for thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}

# theft
# irf
irf_theft_1 <- irf(VAR_thefts,
                   impulse = "residuals",
                   response = "thefts",
                   n.ahead = 15)

# ggplot
irf_theft_1_gg <- data.frame(
  irf_theft_1$irf$residuals[,1],
  irf_theft_1$Lower$residuals[,1],
  irf_theft_1$Upper$residuals[,1]
)

@@

irf_theft_1_plot <- ggplot(irf_theft_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  ggtitle("") +
  xlab("Days after a new COVID-19 case")+
  ylab("Thefts")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

@@

Finally, for the city of Seattle, only burglaries stood out based on the fitted VAR models. We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as larceny ($p=0.25$), vandalism ($p=0.79$), vehicle theft ($p=0.31$) and theft from vehicle ($p=0.46$). With a relaxed 90% confidence level, a statistically significant Granger causality was found between COVID-19 cases impacting burglaries ($p=0.073$). The impulse response of a new case on burglaries and the model predictions are outlined in Figure \@ref(fig:r41).

@@

```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
causality(VAR_larceny, cause = colnames(combined_diff)[1])
causality(VAR_larceny, cause = "residuals")

causality(VAR_burglary, cause = colnames(combined_diff)[2])
causality(VAR_burglary, cause = "residuals")
causality(VAR_vandalism, cause = colnames(combined_diff)[3])
causality(VAR_vandalism, cause = "residuals")
causality(VAR_vehicle_theft, cause = colnames(combined_diff)[4])
causality(VAR_vehicle_theft, cause = "residuals")
causality(VAR_theft_fromvehicle, cause = colnames(combined_diff)[5])
causality(VAR_theft_fromvehicle, cause = "residuals")

@@

```{r r41, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on burglaries (left), and VAR model prediction for burglaries (right). ", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
lags = c(1:16)

@@

irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
  geom_line(aes(y = mean), color = "black") +
  geom_hline(yintercept = 0, color = "blue") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
  theme_classic() +
  xlab("Number of days after a confimed COVID-19 case")+
  ylab("Number of bulglary cases")

@@

Across all of the studied cities, we found that the crimes most significantly associated with COVID-19 cases are vehicle thefts and burglaries/thefts, with vandalism and assault also emerging as significant in some sectors. In the subsequent section we now discuss various implications of these findings and also discuss limitations of the study.
